---
title: "Matching in R"
author: "Katherine Siegel"
date: "2023-02-16"
output: html_document
---

## Description
Code to demonstrate matching in R. Adapted from the supplementary materials from Butsic, V. et al. (2017): Quasi-experimental methods enable stronger inferences from observational data in ecology. (c) Matthias Baumann (2017-01-10).	

In the Butsic et al. paper, they used the example of the impact of wildfire on species richness. Here, we will simulate data with a known treatment effect of fire on species richness. We will then compare the estimated effect we get through a naive ordinary least squares (OLS) regression approach to the effect we estimate when we use matching methods to control for observable confounding variables.

## Set up
Load required packages. In this demo, we will use the package "MatchIt" for the matching process.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

### Install MatchIt if you need to
install.packages("MatchIt",
                 repos = "http://cran.us.r-project.org")

### Packages for the purposes of the quarto website (no need to download these to run the example)
# install.packages(c("jsonlite", "rmarkdown"),
#                  repos = "http://cran.us.r-project.org")

### Load libraries
library(tidyverse)
library(MatchIt)
```

## Simulate data
Simulated data is handy because we know the true effect of the treatment variable. Here, we'll write a function to simulate a dataset where we know the true effect of fire on species richness.
```{r}
### Function to simulate data and write it as a dataframe
simulate_data <- function(){
  
  ### Create variables in a dataframe
  
  ### Make column for observation ID
  df <- data.frame(id = seq(1,1000),
                   
                   ### Add columns for explanatory variables
                   
                   ### Add column for treatment variable
                   fire = c(rep(0,500), rep(1,500)),
                   
                   ### And the rest of the covariates
                   slope = c(runif(500, min = 50, max = 90), 
                             runif(500, min = 65, max = 150)),
                   elevation = c(runif(500, min = 150, max = 185), 
                                 runif(500, min = 165, max = 200)),
                   stream = runif(1000, min = 0, max = 1),
                   
                   ### And the error term
                   error = rnorm(1000, mean = 0, sd = 5))
  
  ### Add a slope*slope variable
  df <- df %>%
    mutate(slope2 = slope^2)
  
  ### Make column for outcome variable (species richness) 
  df <- df %>%
    mutate(species_richness = 1 + 5*fire + 0.07*slope + 0.05*elevation + 2*stream - 0.005*slope2 + error)
  return(df)
}
```
*We know that the true effect of the treatment variable (fire) is a 5x increase in the response variable (species richness).*

## Estimate the effect using ordinary least squares
```{r}
### Write a function to generate data and analyze using OLS 
ols_fun <- function(){
  
  ### Simulate the dataset
  data <- simulate_data()
  
  ### Run OLS regression
  ols <- lm(species_richness ~ fire + slope + elevation + stream, 
            data = data)
  
  ### Extract model coefficients and standard error
  fire_coeff <- coef(summary(ols))["fire", "Estimate"]
  fire_se <- coef(summary(ols))["fire", "Std. Error"]
  list <- list(fire_coeff, fire_se)
}

### Apply the function to 100 replicates
ols_sim <- replicate(100, ols_fun())

### Extract the model estimates
ols_fire_est <- unlist(ols_sim[1, ])

### Print mean, standard deviation, minimum, and maximum values for coefficient estimates
c(mean(ols_fire_est), sd(ols_fire_est), 
  min(ols_fire_est), max(ols_fire_est))

### Extract the standard deviations
ols_fire_sd <- unlist(ols_sim[2, ])

### Print mean, standard deviation, minimum, and maximum standard deviation of coefficient estimates
c(mean(ols_fire_sd), sd(ols_fire_sd), 
  min(ols_fire_sd), max(ols_fire_sd))
```
*The effect estimated by OLS is incorrect-- it should be 5.*

## Use pre-regression matching, then run the regression
```{r}
### Write a function to generate the data, use matching to subset the data, and run a regression on the matched data
pps_fun <- function(){
  
  ### Simulate the dataset
  data <- simulate_data()
  
  ### Match the data on the observed covariates
  match <- matchit(fire ~ slope + 
                     elevation + 
                     stream, 
                   
                   ### set method to use for matching
                   method = "nearest", 
                   
                   ### tell it what data source to draw matches from
                   data = data, 
                   
                   ### tell it to use logistic regression for the matching
                   distance = "glm", 
                   link = "probit",
                   
                   ### specify which order to draw potential points from the full dataset
                   m.order = "random",
                   
                   ### set a maximum distance for the matches
                   caliper = 0.10)
  
  ### Extract the matched data from the full dataset
  matched_data = match.data(match)
  
  ### Run OLS on the matched dataset
  ols <- lm(species_richness ~ fire + slope + 
              elevation + stream, 
            data = matched_data)
  
  ### Extract model coefficients
  fire_coeff <- coef(summary(ols))["fire", "Estimate"]
  fire_se <- coef(summary(ols))["fire", "Std. Error"]
  list <- list(fire_coeff, fire_se) 
}

### Apply the function to 100 replicates
pps_sim <- replicate(100, pps_fun())

### Extract the model estimates
pps_fire_est <- unlist(pps_sim[1,])

### Print mean, standard deviation, minimum, and maximum values for coefficient estimates
c(mean(pps_fire_est), sd(pps_fire_est), 
  min(pps_fire_est), max(pps_fire_est))

### Extract the standard deviations
pps_fire_sd <- unlist(pps_sim[2,])

### Print mean, standard deviation, minimum, and maximum standard deviation of coefficient estimates
c(mean(pps_fire_sd), sd(pps_fire_sd), 
  min(pps_fire_sd), max(pps_fire_sd))
```

## Take a closer look at the matching process
### Pre-matching data
```{r}
### Simulate a dataset
data_for_match <- simulate_data()

### Make fire a factor variable
data_for_match <- data_for_match %>%
  mutate_at(vars(fire), 
            funs(factor))

### Take a look at the balance of the covariates before matching
data_for_match %>%
  dplyr::select(id, fire, slope, elevation, stream) %>%
  gather(variable, value, slope:stream, factor_key = TRUE) %>%
  ggplot(aes(x = variable, y = value, color = fire)) +
  geom_boxplot() +
  xlab("Variable") + ylab("Value")
### fires: 0 = unburned, 1 = burned

### You can also look at it in table form
covariate_summ <- data_for_match %>%
  group_by(fire) %>%
  summarise(slope_mean = mean(slope),
            slope_sd = sd(slope),
            elevation_mean = mean(elevation),
            elevation_sd = sd(elevation),
            stream_mean = mean(stream),
            stream_sd = sd(stream))
```
We can see that there are issues with the balance between the burned and unburned sample units. The burned areas are on steeper slopes and higher elevations, on average.

### Match the data
```{r}
### Match the data on the observed covariates
match <- matchit(fire ~ slope + 
                   elevation + 
                   stream, 
                 
                 ### set method to use for matching
                 method = "nearest", 
                 
                 ### tell it what data source to draw matches from
                 data = data_for_match, 
                 
                 ### tell it to use logistic regression for calculating the propensity scores 
                 distance = "glm", 
                 link = "probit",
                 
                 ### specify which order to draw potential points from the full dataset
                 m.order = "random",
                 
                 ### set a maximum distance for the matches
                 caliper = 0.10)

### Take a look at the quality of the matches
match_quality <- summary(match, 
                         standardize = TRUE)

### Let's see how many points were matched
match_quality_nn <- as.data.frame(match_quality$nn)

### Let's look at the pre-match covariate balance
match_quality_unmatched <- as.data.frame(match_quality$sum.all)

### What does the covariate balance look like after matching?
match_quality_summary <- as.data.frame(match_quality$sum.matched)
### Ideally, you want the standardized mean differences in the matched dataset to be < 0.25 (reference: Schleicher et al. 2020. Statistical matching for conservation science. Conserv. Biol. 34:538–549. https://doi.org/10.1111/cobi.13448).

### You can also look at the amount of bias *reduction achieved through matching
# match_quality_reduction <- as.data.frame(match_quality$reduction)
```

#### Visualizing the match quality
```{r}
### You can also use a fun interactive command to visualize the pre- and post-match covariate spread
# plot(match, interactive = FALSE)

### And you can compare the propensity scores visually
# plot(match, type = "jitter", interactive = FALSE)
```

### Analyzing the matched subset of data
```{r}
### First, extract the matches from the full dataset
demo_matched <- match.data(match)

### Then run the regression, including the covariates in the model
fit_matched <- lm(species_richness ~ fire + slope + 
                    elevation + stream, 
                  data = demo_matched)
```



